Partitioned Local Depth (PaLD) Community Analyses in R

### 2023-09-28 2023-r-journal/dagostino-mcgowan.tex
upload files

\abstract{%
Partitioned Local Depth (PaLD) is a framework for a holistic
consideration of the community structure of distance-based data. This
paper describes an R package, \CRANpkg{pald}, for calculating
Partitioned Local Depth (PaLD) probabilities, implementing community
analyses, and creating data visualizations to display community
structure. We describe how to use the package, walk through several
examples, and contrast the results with commonly used techniques.
}

@@

Throughout this paper, we will display the network obtained from
cohesion using a force-directed graph drawing algorithm and emphasize
the strong ties (colored by connected component). We refer to the
connected components of the network of strong ties as community
``clusters.'' Note that to qualify as a cluster in this definition, one
may not have any strong ties with those outside the cluster, and thus
the existence of disjoint groups is a strong signal for separation.
Here, clusters are identified without additional user inputs nor
optimization criteria. If one wishes to further break the community
graph into groups, one may consider using community detection methods
(such as spectral clustering or the Louvain algorithm), as available,
say, in the \CRANpkg{igraph} package. Though only briefly considered
here, one may also use the collection of strong ties in place of
(weighted) k-nearest neighbors in settings such as classification and
smoothing. Overall, the structural information obtained from local
depth, cohesion and community graphs can provide a holistic perspective
on the data which does not require the use of distributional
assumptions, optimization criteria nor additional user inputs.

We present a new package, \CRANpkg{pald}, for calculating Partitioned
Local Depth (PaLD) probabilities, implementing community analyses, and
creating data visualizations to display community structure. This paper
will describe how to use the package, walk through several examples, and
compare the method to commonly used techniques in R. Together, these
demonstrate both the novelty of the method and utility of the
implementation in package described.

@@

The input for the Partitioned Local Depths (PaLD) computations is a
distance matrix or \texttt{dist} object. Note that the collection of
input distances (or dissimilarities) does not need to satisfy the
triangle inequality nor be symmetric.

For demonstration purposes, we will show how one can compute a distance
matrix from an input data frame with, say, two variables \texttt{x1} and
\texttt{x2}. The input data may be of any dimension; in fact the PaLD
framework provides advantages when considering high-dimensional data
(see the \textbf{Examples} section as well as
\citet{berenhaut2022social}).

@@

The \texttt{dist()} function converts an input data frame into a
distance matrix, as demonstrated below. If the data are already provided
as a distance matrix (or \texttt{dist} object), the user can skip to the
next step. Note that the distance matrix needed for the subsequent
functions doesn't need to be a \texttt{dist} object and \emph{need not}
be symmetric.

@@

The function above creates a \texttt{dist} object. If converted to a
matrix, this will be an \(n\times n\) distance matrix, where \(n\)
corresponds to the number of observations in the original data frame, in
this example \(n = 6\).

This \texttt{dist} object, or a distance matrix, can then be passed to
the \texttt{cohesion\_matrix()} function in order to calculate the
pairwise cohesion values.

Cohesion is an interpretable probability that reflects the strength of
alignment of two points within local regions. It captures aspects of the
relative positioning of points and accounts for varying density across
the space.

@@

The \emph{cohesion matrix} output by the \texttt{cohesion\_matrix()}
function is the main input for the majority of the remaining functions.

@@

The \texttt{local\_depths()} function calculates the \emph{depth} of
each point, outputting a vector of local depths. Local depth is an
interpretable probability which reflects aspects of relative position
and centrality via distance comparisons (i.e., \(d(z, x) < d(z, y)\),
see \citet{berenhaut2022social}).

@@

The \texttt{strong\_threshold()} function will calculate the cohesion
threshold for strong ties. This is equal to half the average of the
diagonal of the cohesion matrix \citep{berenhaut2022social}, and is a
threshold that may be used to distinguish between strong and weak ties.

@@

The function \texttt{cohesion\_strong()} will update the cohesion matrix
to set all weak ties to zero (via the \texttt{strong\_threshold()}
function). Optionally, the matrix will also be symmetrized, using the
entry-wise (parallel) minimum of the cohesion matrix and its transpose,
with the default parameter \texttt{symmetric\ =\ TRUE}.

@@

The \texttt{community\_graphs()} function takes the cohesion matrix and
creates \CRANpkg{igraph} objects, graphs that describe the relationship
between the points. This function will output a list of three objects:

\begin{itemize}
\tightlist
\item
  \texttt{G}: the weighted (community) graph whose edge weights are
  mutual cohesion
\item
  \texttt{G\_strong}: the weighted (community) graph consisting of edges
  for which mutual (symmetrized) cohesion (i.e.~the minimum of the two
  directed cohesion values for any given pair) is greater than the
  threshold for strong ties
\item
  \texttt{layout}: the graph layout, using the Fruchterman Reingold (FR)
  force-directed graph drawing for the graph \texttt{G}
\end{itemize}

\begin{Schunk}
\begin{Sinput}
graphs <- community_graphs(cohesion)
graphs[["G_strong"]]
\end{Sinput}
\begin{Soutput}
#> IGRAPH 4d37a4f UNW- 6 3 -- 
#> + attr: name (v/c), weight (e/n)
#> + edges from 4d37a4f (vertex names):
#> [1] A--B A--E D--F
\end{Soutput}
\end{Schunk}

Here we see that there are three connected components, ties \texttt{A-B}
and \texttt{A-E} which form the first cluster, and the tie \texttt{D-F}
which forms another.

The \texttt{any\_isolated()} function will check whether there are any
isolated points.

@@

The ``community clusters'' identified by PaLD are the connected
components of the graph of strong ties, \texttt{G\_strong}. To directly
calculate them, we can use the \texttt{community\_clusters()} function.
This will output a data frame with two columns, the first will
correspond to the \texttt{point}, as identified by the row name of the
original input data frame, \texttt{df}, the second will identify the
\texttt{cluster} that each point belongs to.

\begin{Schunk}
\begin{Sinput}
community_clusters(cohesion)
\end{Sinput}
\begin{Soutput}
#>   point cluster
#> A     A       1
#> B     B       1
#> C     C       2
#> D     D       3
#> E     E       1
#> F     F       3
\end{Soutput}
\end{Schunk}

In this example, three clusters are identified with these six points.
Points \texttt{A}, \texttt{B}, and \texttt{E} fall into Cluster 1. Point
\texttt{C} is in Cluster 2 (a cluster of size 1) and points \texttt{D}
and \texttt{F} fall into Cluster 3.

@@

The final category of function is that for data visualization. We can
begin by visualizing the points in the data frame \texttt{df} (Figure
\ref{fig:fig1}). When visualizing these points, it is important to have
the aspect ratio of the x and y axes equal to 1 so as to not distort
distances. When using the \CRANpkg{ggplot2} package for this
visualization, one can use the \texttt{coord\_fixed(ratio\ =\ 1)}
function. If using the \texttt{plot()} function included in the base
library, one can use the \texttt{asp\ =\ 1} argument.

@@

We can pass the cohesion matrix to the
\texttt{plot\_community\_graphs()} function to view the relationship
between points (Figure \ref{fig:fig2}). Notice in this plot the layout
does not match that of the original data frame as seen in Figure
\ref{fig:fig1}. Since our original data is two dimensional, it may be
reasonable to use this as the layout. Figure \ref{fig:fig3} will make
this update as well as update some of the aesthetics, such as employing
more readable labels.

\begin{Schunk}
\begin{Sinput}
plot_community_graphs(cohesion)
\end{Sinput}
\begin{figure}
\includegraphics{dagostino-mcgowan_files/figure-latex/fig2-1} \caption[PaLD graph displaying the relationship between the points in data frame `df`]{PaLD graph displaying the relationship between the points in data frame `df`}\label{fig:fig2}
\end{figure}
\end{Schunk}

The \texttt{layout} argument allows the user to pass a matrix to dictate
the 2-dimensional layout of the graph. For example, if we wanted the
graph to match the visualization displayed in Figure \ref{fig:fig1}, we
could pass \texttt{as.matrix(df)}, or a matrix of the data frame
\texttt{df} to the \texttt{layout} argument (Figure \ref{fig:fig3}).
Additionally, the \texttt{plot\_community\_graphs()} function will also
permit parameters that can be passed to the \texttt{plot.igraph()}
function. We can pass arguments to the \texttt{plot.igraph} function via
the \texttt{...} argument; for example to increase the vertex size and
change the vertex label color, we can specify
\texttt{vertex.size\ =\ 100} and
\texttt{vertex.label.color\ =\ "white"}. Additionally, to allow axes, we
use \texttt{axes\ =\ TRUE}, and to put them back on the original scale
we set \texttt{rescale\ =\ FALSE}, resetting the axis limits using
\texttt{xlim} and \texttt{ylim}. The \texttt{par(pty\ =\ "s")} function
forces the subsequent plot to be square.

@@

We will demonstrate the utility of the \CRANpkg{pald} package in three
illustrative examples.

\hypertarget{clustering-tissue-gene-expression-data}{%
\subsubsection{Clustering tissue gene expression
data}\label{clustering-tissue-gene-expression-data}}

The first example is from a subset of data from \citet{zilliox2007gene},
\citet{mccall2011gene}, and \citet{mccall2014gene}, obtained from the
\textbf{tissuesGeneExpression} bioconductor package \citep{tissue}
consisting of 22,215-dimensional gene expression data from 189 tissue
samples. A \texttt{dist} object was created using this data set and is
included in the \CRANpkg{pald} package in an object called
\texttt{tissue\_dist}.

@@

The \texttt{community\_clusters()} function can be used to identify the
community cluster corresponding to each tissue sample. Since the output
is a data frame, we can summarize the result using commonly employed
data analysis techniques. For demonstration purposes, we will use the
\CRANpkg{dplyr} package to summarize the contribution of clusters.

\begin{Schunk}
\begin{Sinput}
community_clusters(tissue_cohesion) |>
  dplyr::count(cluster, point)
\end{Sinput}
\begin{Soutput}
#>    cluster       point  n
#> 1        1 endometrium 15
#> 2        1      kidney 39
#> 3        2 hippocampus 31
#> 4        3  cerebellum 26
#> 5        4  cerebellum  1
#> 6        5       colon 33
#> 7        6       colon  1
#> 8        7       liver  7
#> 9        8  cerebellum  1
#> 10       9       liver 17
#> 11      10  cerebellum  2
#> 12      11       liver  2
#> 13      12  cerebellum  1
#> 14      13  cerebellum  4
#> 15      14  cerebellum  2
#> 16      15  cerebellum  1
#> 17      16    placenta  2
#> 18      17    placenta  1
#> 19      18    placenta  3
\end{Soutput}
\end{Schunk}

From this, we can glean that Cluster 1 consists of two types of tissue,
the kidney and endometrium. Cluster 2 is comprised of only the
hippocampus.

We can also display the relationships between tissue samples using the
\texttt{plot\_community\_graphs()} function (Figure \ref{fig:fig4}). For
clarity of the display, we show how to remove the labels using
\texttt{show\_labels\ =\ FALSE}. We will instead color by the labels by
passing these to the \texttt{vertex.color} parameter to the
\texttt{plot.igraph} function (via the \texttt{...} argument).
Similarly, we can add a legend using the \texttt{legend()} function, as
you would for an \CRANpkg{igraph} visualization. Additionally, we use
the \texttt{edge\_width\_factor} and \texttt{emph\_strong} arguments to
adjust the width of the lines between and within PaLD clusters.

@@

\begin{Schunk}
\begin{figure}
\includegraphics[width=1\linewidth]{fig5} \caption[Community cluster network for the tissue data]{Community cluster network for the tissue data. The line colors indicate the PaLD clusters, the point colors indicate the tissue classification.}\label{fig:fig4}
\end{figure}
\end{Schunk}

@@

This example performs a PaLD analysis on a data set from \citet{dyen92}
that examines the relationship between 87 Indo-European languages from
the perspective of cognates, coded using 2,655-dimensional binary
vectors. A \texttt{dist} object was created from this data set and is
included in the \CRANpkg{pald} package in an object called
\texttt{cognate\_dist}.

This example will demonstrate how you can apply functions in the
\CRANpkg{igraph} package to objects output from the \CRANpkg{pald}
package. We can first use the \texttt{cohesion\_matrix()} function to
calculate the cohesion matrix and the \texttt{community\_graphs()}
function to create a list with the weighted community graph, the
weighted community graph with only strong ties included, and the layout.
From this, we can extract the graph with only the strong ties, here
called \texttt{cognate\_graph\_strong}.

@@

We can then use the \texttt{neighbors()} function from the
\CRANpkg{igraph} package to extract the strong neighbors in this graph.
For example, if we wanted to extract all neighbors for the language
``French'', we could run the following.

\begin{Schunk}
\begin{Sinput}
french_neighbors <- igraph::neighbors(cognate_graph_strong, "French")
french_neighbors
\end{Sinput}
\begin{Soutput}
#> + 8/87 vertices, named, from a461d26:
#> [1] Italian         Ladin           Provencal       Walloon        
#> [5] French_Creole_C French_Creole_D Spanish         Catalan
\end{Soutput}
\end{Schunk}

@@

We can again use the \texttt{plot\_community\_graphs()} function to
visualize the community clusters (Figure \ref{fig:figlang}). One may
note the commonly identifiable language clusters and that, under a
slight rotation, some of the underlying geography is mirrored in the
plot.

@@

\begin{itemize}
\tightlist
\item
  \texttt{exdata1} is a data set consisting of 8 points to recreate
  Figure 1 in \citet{berenhaut2022social}
\item
  \texttt{exdata2} is a data set consisting of 16 points to recreate
  Figure 2 in \citet{berenhaut2022social}
\item
  \texttt{exdata3} is a data set consisting of 240 points to recreate
  Figure 4D in \citet{berenhaut2022social}
\end{itemize}

@@

We will demonstrate how we can compare PaLD to two clustering methods:
\emph{k}-means and hierarchical clustering. The code below calculates
the cohesion matrix (\texttt{exdata\_cohesion}) as well as the clusters
via PaLD (\texttt{exdata\_pald}), \emph{k}-means
(\texttt{exdata\_kmeans}) and hierarchical clustering using complete
linkage (\texttt{exdata\_hclust}).

@@

We can compare this to the clustering generated by \emph{k}-means and
hierarchical clustering (Figure \ref{fig:fig5}).

\begin{Schunk}
\begin{Sinput}
par(mfrow = c(1, 3), pty = "s")
plot(
  exdata3,
  pch = 16,
  col = pald_colors[exdata_pald],
  xlab = "",
  ylab = "",
  main = "PaLD Clusters",
  asp = 1
)
plot(
  exdata3,
  pch = 16,
  col = pald_colors[exdata_kmeans],
  xlab = "",
  ylab = "",
  main = "K-Means Clusters (k = 8)",
  asp = 1
)
plot(
  exdata3,
  pch = 16,
  col = pald_colors[exdata_hclust],
  xlab = "",
  ylab = "",
  main = "Hiearchical Clusters (k = 8)",
  asp = 1
)
\end{Sinput}
\begin{figure}
\includegraphics{dagostino-mcgowan_files/figure-latex/fig5-1} \caption[PaLD clustering of randomly generated example data (from Figure 4D from Berenhaut et al]{PaLD clustering of randomly generated example data (from Figure 4D from Berenhaut et al. (2022)) compared to k-means and hierarchical clustering with k = 8.}\label{fig:fig5}
\end{figure}
\end{Schunk}

Cohesion is particularly useful when considering data with varying local
density; see \citet{berenhaut2022social} for further examples,
discussion, and theoretical results. Note that the PaLD algorithm is
able to detect the eight natural groups within the data without the use
of any additional inputs (e.g., number of clusters) nor optimization
criteria. Despite the user input of the ``correct'' number of clusters
(i.e., \(k = 8\)) both \emph{k}-means and hierarchical clustering did
not give the desired result.

@@

\textbf{TODO intro to data here} \citep{muthukrishna2020beyond}

First, we will create the cohesion matrix using the \texttt{cultures}
data, provided in \CRANpkg{pald}, and proceed to plot the community
structure.

@@

This example demonstrates key features of the PaLD approach that
distinguish it from commonly used clustering techniques. In addition to
viewing the community structure as seen in Figure \ref{fig:figculture},
we can also visualize a two dimensional plot of cohesion versus distance
for each of the included distance pairs using the
\texttt{dist\_cohesion\_plot()} function, as seen below (Figure
\ref{fig:figco}).

\begin{Schunk}
\begin{Sinput}
dist_cohesion_plot(cultures)
\end{Sinput}
\begin{figure}
\includegraphics{dagostino-mcgowan_files/figure-latex/figco-1} \caption[Distance versus cohesion for each TODO The identified communties are shown via the 5 colors -- these map to the same clusters idenfitied in Figure 7]{Distance versus cohesion for each TODO The identified communties are shown via the 5 colors -- these map to the same clusters idenfitied in Figure 7.}\label{fig:figco}
\end{figure}
\end{Schunk}

>>>>

\abstract{%
Partitioned Local Depth (PaLD) is a framework for holistic consideration
of community structure for distance-based data. This paper describes an
R package, \CRANpkg{pald}, for calculating Partitioned Local Depth
(PaLD) probabilities, implementing community analyses, and creating data
visualizations to display community structure. We describe how to use
the package, walk through several examples, and discuss the approach in
relation to commonly used techniques.
}

@@

Networks obtained from cohesion can be displayed using a force-directed
graph drawing algorithm; here, we will graphically emphasize strong ties
(colored by connected component). We refer to the connected components
of the network of strong ties as community ``clusters.'' Note that to
qualify as a cluster in this definition, one may not have any strong
ties with those outside the cluster, and thus the existence of disjoint
groups is a strong signal for separation. Here, clusters are identified
without additional user inputs nor optimization criteria. If one wishes
to further break the community graph into groups, one may consider using
community detection methods (such as spectral clustering or the Louvain
algorithm), as available, say, in the \CRANpkg{igraph} package. One may
also use the collection of strong ties in place of (weighted)
\(k\)-nearest neighbors in settings such as classification and
smoothing. Overall, the structural information obtained from local
depth, cohesion and community graphs can provide a holistic perspective
to the data which does not require the use of distributional
assumptions, optimization criteria nor additional user inputs.

We present a new package, \CRANpkg{pald}, for calculating Partitioned
Local Depth (PaLD) probabilities, implementing community analyses, and
creating data visualizations to display community structure. This paper
describes how to use the package, walks through several examples, and
discusses the approach in the context of commonly used techniques,
demonstrating both the novelty of the method and utility of the
implementation in the package described.

@@

For the purposes of the \CRANpkg{pald} package, the sole input for the
Partitioned Local Depth (PaLD) computations is a distance matrix or
\texttt{dist} object. Note that the collection of input distances (or
dissimilarities) does not need to satisfy the triangle inequality nor be
symmetric. More generally, the method only requires triplet distance
comparisons, as opposed to exact numeric distances (see
\citet{berenhaut2022social}).

For demonstration purposes, we first show how one can compute a distance
matrix from an input data frame with, say, two variables \texttt{x1} and
\texttt{x2}. The input data may be of any dimension; in fact the PaLD
framework provides advantages when considering high-dimensional data
(see the \textbf{Examples} section as well as
\citet{berenhaut2022social}).

@@

The \texttt{dist} function returns a (default Euclidean) pairwise
distance matrix for an input data frame, as demonstrated below. If the
data are already provided as a distance matrix (or \texttt{dist}
object), the user can skip to the next step. Note that the distance
matrix needed for the subsequent functions does not need to be a
\texttt{dist} object and \emph{need not} be symmetric.

@@

The function above creates a \texttt{dist} object. If converted to a
matrix, this will be an \(n\times n\) distance matrix, where \(n\)
corresponds to the number of observations in the original data frame (in
this example \(n = 6\)).

This \texttt{dist} object, or a distance matrix, can then be passed to
the \texttt{cohesion\_matrix} function in order to calculate pairwise
cohesion values.

Cohesion is an interpretable probability that reflects the strength of
alignment of two points within local regions. It captures aspects of the
relative positioning of points and accounts for varying density across
the space (see \citet{berenhaut2022social}).

@@

The \emph{cohesion matrix} output by the \texttt{cohesion\_matrix}
function is the main input for the majority of the remaining functions.

@@

The \texttt{local\_depths} function calculates the \emph{depth} of each
point, outputting a vector of local depths. Local depth is an
interpretable probability which reflects aspects of relative position
and centrality via distance comparisons (e.g., \(d(z, x) < d(z, y)\),
see \citet{berenhaut2022social}).

@@

The \texttt{strong\_threshold} function will calculate the cohesion
threshold for strong ties, which reflects typical cohesion for local
points (see \citet{berenhaut2022social}). Computationally, this is equal
to half the average of the diagonal of the cohesion matrix, and is a
threshold that may be used to distinguish between strong and weak ties.

@@

The function \texttt{cohesion\_strong} will update the cohesion matrix
to set all weak ties to zero (via the \texttt{strong\_threshold}
function). Optionally, the matrix will also be symmetrized, using the
entry-wise (parallel) minimum of the cohesion matrix and its transpose,
with the default parameter \texttt{symmetric\ =\ TRUE}.

@@

The \texttt{community\_graphs} function takes the cohesion matrix and
creates \CRANpkg{igraph} objects, graphs that describe the (symmetrized)
relationship structure between points. This function will output a list
of three objects:

\begin{itemize}
\tightlist
\item
  \texttt{G}: the weighted (community) graph whose edge weights are
  mutual cohesion
\item
  \texttt{G\_strong}: the weighted (community) graph consisting of edges
  for which mutual (symmetrized) cohesion (i.e.~the minimum of the two
  directed cohesion values for any given pair) is greater than the
  threshold for strong ties
\item
  \texttt{layout}: the graph layout. By default this is provided by the
  Fruchterman Reingold (FR) force-directed graph drawing algorithm for
  the graph \texttt{G}, as implemented in the \CRANpkg{igraph} package.
\end{itemize}

\begin{Schunk}
\begin{Sinput}
graphs <- community_graphs(cohesion)
graphs[["G_strong"]]
\end{Sinput}
\begin{Soutput}
#> IGRAPH 198a54e UNW- 6 3 -- 
#> + attr: name (v/c), weight (e/n)
#> + edges from 198a54e (vertex names):
#> [1] A--B A--E D--F
\end{Soutput}
\end{Schunk}

Here we see that there are three connected components, ties \texttt{A-B}
and \texttt{A-E} form the first community cluster, and the tie
\texttt{D-F} which forms another.

The \texttt{any\_isolated()} function will check whether there are any
isolated points (according to cohesion).

@@

The ``community clusters'' identified by PaLD are the connected
components of the graph of strong ties, \texttt{G\_strong}. To directly
calculate these, we can use the \texttt{community\_clusters} function.
This will output a data frame with two columns; the first corresponds to
the \texttt{point}, as identified by the row name of the original input
data frame, \texttt{df}, the second identifies the \texttt{community}
that the point belongs to.

\begin{Schunk}
\begin{Sinput}
community_clusters(cohesion)
\end{Sinput}
\begin{Soutput}
#>   point community
#> A     A         1
#> B     B         1
#> C     C         2
#> D     D         3
#> E     E         1
#> F     F         3
\end{Soutput}
\end{Schunk}

In this example, three communities are identified with these six points.
Points \texttt{A}, \texttt{B}, and \texttt{E} fall into Community 1.
Point \texttt{C} is in Community 2 (a community of size 1) and points
\texttt{D} and \texttt{F} fall into Community 3.

@@

The final category of function is that for data visualization. We can
begin by visualizing the points in the data frame \texttt{df} (Figure
\ref{fig:fig1}). When visualizing these points, it is important to have
the aspect ratio of the x and y axes equal to 1 so as to not distort
distances. When using the \CRANpkg{ggplot2} package for this
visualization, one can use the command
\texttt{coord\_fixed(ratio\ =\ 1)}. If using the \texttt{plot} function
included in the base library, one can use the \texttt{asp\ =\ 1}
argument.

@@

We can pass the cohesion matrix to the \texttt{plot\_community\_graphs}
function to view the relationship between points (Figure
\ref{fig:fig2}). The function will also permit parameters that can be
passed to \texttt{plot.igraph} via the \texttt{...} argument.

\begin{Schunk}
\begin{Sinput}
plot_community_graphs(cohesion, 
                      vertex.label.cex = 2,
                      vertex.label.dist = 0.9)
\end{Sinput}
\begin{figure}
\includegraphics{dagostino-mcgowan_files/figure-latex/fig2-1} \caption[PaLD graph displaying the relationship between the points in data frame `df`]{PaLD graph displaying the relationship between the points in data frame `df`}\label{fig:fig2}
\end{figure}
\end{Schunk}

Notice in this plot the force-directed layout does not match that of the
original data frame as seen in Figure \ref{fig:fig1}. Since our original
data is two-dimensional, it may be reasonable to use the latter as the
layout for plotting. Figure \ref{fig:fig3} includes this update as well
as others addressing some of the aesthetics, such as employing more
readable labels. The \texttt{layout} argument allows the user to pass a
matrix to dictate the 2-dimensional layout of the graph. For example, if
we wanted the graph to match the visualization displayed in Figure
\ref{fig:fig1}, we could pass \texttt{as.matrix(df)} (a matrix of the
data frame \texttt{df}) to the \texttt{layout} argument (Figure
\ref{fig:fig3}). Here, we increase the vertex size and change the vertex
label color, through the argument specifications
\texttt{vertex.size\ =\ 100} and
\texttt{vertex.label.color\ =\ "white"}. Additionally, to allow axes, we
use \texttt{axes\ =\ TRUE}, and to put these back on the original scale
we set \texttt{rescale\ =\ FALSE}, resetting the axis limits using
\texttt{xlim} and \texttt{ylim}. The \texttt{par(pty\ =\ "s")} function
forces the subsequent plot to be square.

@@

We will demonstrate the utility of the \CRANpkg{pald} package through
several illustrative examples.

\hypertarget{community-analysis-for-tissue-gene-expression-data}{%
\subsubsection{Community analysis for tissue gene expression
data}\label{community-analysis-for-tissue-gene-expression-data}}

The first example is from a subset of data from \citet{zilliox2007gene},
\citet{mccall2011gene}, and \citet{mccall2014gene}, obtained from the
\textbf{tissuesGeneExpression} bioconductor package \citep{tissue}
consisting of 22,215-dimensional gene expression data from 189 tissue
samples. A (Euclidean) \texttt{dist} object was created using this data
set and is included in the \CRANpkg{pald} package in an object called
\texttt{tissue\_dist}.

@@

The \texttt{community\_clusters} function can be used to identify the
community corresponding to each tissue sample. Since the output is a
data frame, we can summarize the result using commonly employed R
functions. For demonstration purposes, we will use the \CRANpkg{dplyr}
package to summarize the community contributions.

\begin{Schunk}
\begin{Sinput}
community_clusters(tissue_cohesion) |>
  dplyr::count(community, point)
\end{Sinput}
\begin{Soutput}
#>    community       point  n
#> 1          1 endometrium 15
#> 2          1      kidney 39
#> 3          2 hippocampus 31
#> 4          3  cerebellum 26
#> 5          4  cerebellum  1
#> 6          5       colon 33
#> 7          6       colon  1
#> 8          7       liver  7
#> 9          8  cerebellum  1
#> 10         9       liver 17
#> 11        10  cerebellum  2
#> 12        11       liver  2
#> 13        12  cerebellum  1
#> 14        13  cerebellum  4
#> 15        14  cerebellum  2
#> 16        15  cerebellum  1
#> 17        16    placenta  2
#> 18        17    placenta  1
#> 19        18    placenta  3
\end{Soutput}
\end{Schunk}

From this, we can glean, for instance, that Community 1 consists of two
types of tissue, the kidney and endometrium, Community 2 is comprised of
only the hippocampus, etc. More informatively, we can display
relationships between tissue samples, both locally and globally through
the \texttt{plot\_community\_graphs} function (Figure \ref{fig:fig4}).
For clarity of the display, we show how to remove the labels using
\texttt{show\_labels\ =\ FALSE}. We will instead color according to the
labels by passing these to the \texttt{vertex.color} argument for the
\texttt{plot.igraph} function (via the \texttt{...} argument).
Similarly, we can add a legend using the \texttt{legend()} function, as
you would for an \CRANpkg{igraph} visualization. Additionally, we use
the \texttt{edge\_width\_factor} and \texttt{emph\_strong} arguments to
adjust the width of the lines between and within PaLD communities.

@@

\begin{Schunk}
\begin{figure}
\includegraphics[width=1\linewidth]{fig5} \caption[Community cluster network for the tissue data]{Community cluster network for the tissue data. The line colors indicate the PaLD communities; the point colors indicate the tissue classification.}\label{fig:fig4}
\end{figure}
\end{Schunk}

Note that while Community 1 does contain kidney and endometrium tissues,
the respective sub-communities are well-separated. Of the 357 strong
inter-community ties within Community 1, only three are between the two
sub-communities. In fact, these are the only strong ties between
different tissue types in the community cluster network (of 1064 total);
a summary of community strong ties can be produced via the following
code. Note that \texttt{tissue\_graph\_strong} is an \CRANpkg{igraph}
object corresponding to the graph displayed in Figure \ref{fig:fig4},
restricted only to strong ties.

\begin{Schunk}
\begin{Sinput}
tissue_graphs <- community_graphs(tissue_cohesion)
tissue_graph_strong <- tissue_graphs[["G_strong"]]
E <- igraph::get.edgelist(tissue_graph_strong)
table(apply(E, 1, paste, collapse = ","))
\end{Sinput}
\begin{Soutput}
#> 
#>   cerebellum,cerebellum             colon,colon endometrium,endometrium 
#>                     186                     241                      81 
#> hippocampus,hippocampus      kidney,endometrium           kidney,kidney 
#>                     191                       3                     273 
#>             liver,liver       placenta,placenta 
#>                      85                       4
\end{Soutput}
\end{Schunk}

@@

This example explors a data set from \citet{dyen92} that summarizes
relationships between 87 Indo-European languages from the perspective of
cognates, coded using 2,655-dimensional binary vectors. A \texttt{dist}
object was created from this data set and is included in the
\CRANpkg{pald} package in an object called \texttt{cognate\_dist}.

Here we will demonstrate how one can further apply functions in the
\CRANpkg{igraph} package to objects output from the \CRANpkg{pald}
package. We can first use the \texttt{cohesion\_matrix} function to
calculate the cohesion matrix and the \texttt{community\_graphs}
function to create a list with the weighted community graph, the
weighted community graph with only strong ties included, and the layout.
From this, we can extract the graph with only the strong ties, here
called \texttt{cognate\_graph\_strong}.

@@

We can then use the \texttt{neighbors} function from the
\CRANpkg{igraph} package to extract the strong neighbors in this graph.
For example, we can extract all neighbors for the language ``French'',
via the followign code.

\begin{Schunk}
\begin{Sinput}
french_neighbors <- igraph::neighbors(cognate_graph_strong, "French")
french_neighbors
\end{Sinput}
\begin{Soutput}
#> + 8/87 vertices, named, from dc83183:
#> [1] Italian         Ladin           Provencal       Walloon        
#> [5] French_Creole_C French_Creole_D Spanish         Catalan
\end{Soutput}
\end{Schunk}

@@

We can again use the \texttt{plot\_community\_graphs} function to
visualize the community clusters (Figure \ref{fig:figlang}). One may
note the commonly identifiable language clusters and that, under a
slight rotation, some of the underlying geography is mirrored in the
plot.

@@

\begin{itemize}
\tightlist
\item
  \texttt{exdata1} is a data set consisting of 8 points used to create
  Figure 1 in \citet{berenhaut2022social}
\item
  \texttt{exdata2} is a data set consisting of 16 points used to create
  Figure 2 in \citet{berenhaut2022social}
\item
  \texttt{exdata3} is a data set consisting of 240 points used to create
  Figure 4D in \citet{berenhaut2022social}
\end{itemize}

@@

We will contrast community analysis via PaLD to standard culuster
analysis by considering two common clustering methods. The code below
calculates the cohesion matrix (\texttt{exdata\_cohesion}) as well as
the community clusters obtained via PaLD (\texttt{exdata\_pald}), along
with clusters obtained via \emph{k}-means (\texttt{exdata\_kmeans}) and
hierarchical clustering using complete linkage
(\texttt{exdata\_hclust}).

@@

The information is displayed in Figure \ref{fig:fig5}.

\begin{Schunk}
\begin{Sinput}
par(mfrow = c(1, 3), pty = "s")
plot(
  exdata3,
  pch = 16,
  col = pald_colors[exdata_pald],
  xlab = "",
  ylab = "",
  main = "PaLD Communities",
  asp = 1
)
plot(
  exdata3,
  pch = 16,
  col = pald_colors[exdata_kmeans],
  xlab = "",
  ylab = "",
  main = "K-Means Clusters (k = 8)",
  asp = 1
)
plot(
  exdata3,
  pch = 16,
  col = pald_colors[exdata_hclust],
  xlab = "",
  ylab = "",
  main = "Hiearchical Clusters (k = 8)",
  asp = 1
)
\end{Sinput}
\begin{figure}
\includegraphics{dagostino-mcgowan_files/figure-latex/fig5-1} \caption[PaLD, k-means, and hierarchical 8-clustering of randomly generated example data (from \citet{berenhaut2022social}]{PaLD, k-means, and hierarchical 8-clustering of randomly generated example data (from \citet{berenhaut2022social}; Figure 4D).}\label{fig:fig5}
\end{figure}
\end{Schunk}

Cohesion is particularly useful when considering data with varying local
density; see \citet{berenhaut2022social} for further examples,
discussion, and theoretical results. Note that the PaLD algorithm is
able to detect the eight natural groups within the data (along with
inter- and intra-community structure not displayed here) without the use
of any additional inputs (e.g., number of clusters) nor optimization
criteria. Despite the user input of the ``correct'' number of clusters
(i.e., \(k = 8\)) both \emph{k}-means and hierarchical clustering do not
provide the desired result.

@@

In this example we perform a PaLD analysis for cultural distances
obtained in \citet{muthukrishna2020beyond} from two recent waves of the
World Values Survey (2005 to 2009 and 2010 to 2014; see
\citet{inglehart2014world}). Distances are computed using the cultural
fixation index (CFST), which is a measure built on the framework of
fixation indices from population biology
(\citet{bell2009culture,cavalli1994history}). Recall that the foundation
of PaLD in within-triplet comparisons allows for the employment of
application-dependent and non-Euclidean measures of dissimilarity. The
\texttt{dist} object is included in the \CRANpkg{pald} package
(\texttt{cultures}). We will first create the cohesion matrix using the
\texttt{dist} object \texttt{cultures}, and proceed to plot the
community graph.

@@

In addition to viewing the local and global community structure as seen
in Figure \ref{fig:figculture}, the \CRANpkg{pald} package allows for a
two-dimensional display of cohesion against distance for the data, via
the \texttt{dist\_cohesion\_plot} function, as seen below (Figure
\ref{fig:figco}).

\begin{Schunk}
\begin{Sinput}
dist_cohesion_plot(cultures, mutual = TRUE)
\end{Sinput}
\begin{figure}
\includegraphics{dagostino-mcgowan_files/figure-latex/figco-1} \caption{A plot of cohesion versus distance for the data. The identified communities are colored as in Figure \ref{fig:figculture}.}\label{fig:figco}
\end{figure}
\end{Schunk}

Notice here that the magnitude of the distances within each of the
identified communities varies substantially between regions; in fact,
the most disparate two regions in the United States (at distance
\(\approx\) 0.027) are far closer than the two most similar in India (at
distance \(\approx\) 0.043). Despite this, India remains a cohesive
whole, and locally disparate regions in the United States such as East
South Central and California are not strongly cohesive. For discussion
of subtleties in local density (see \citet{berenhaut2022social}.
Additionally, currently available techniques require specification of
parameters, as seen in the previous section. For further discussion of
the cultural distance data in relation to community analysis see
\citet{berenhaut2022social}.
